Speeding up prediction performance of the BDT-based models

### 2017-11-09 acat_paper/paper.tex
the rest

For the most of the ML algorithms, improving the quality of the model usually leads to a higher levels of resource consumption. For example, with the use tree-based boosting algorithms, the increase of the number of trees from tens to thousands usually leads to the significant increase in the quality and at the same time the model size and inference time also amplify. The higher demands for the hardware resources for the more accurate models are usually solved by various technical optimizations model storage mechanism. For example vectoriation of the tree weights storage can be used. Another approach is the modification of the learning process that makes it possible to obtain models with certain properties that make it possible to signigicantly accelerate their application without maintaining the required level of quality.

In this paper, we consider the modification of the learning process of predictive models by the boost decision tree-based algorithm CatBoost [1]. Such modification will accelerate the prediction time of the model while maintaining the required level of the quality.

@@

CatBoost is the implementation of a tree-based gradient boosting in which so-caled oblivious decision trees are used. In this algorithm the values of each feature $i$ are discretized into baskets using the boundaries $B_i$. The boundaries themselves are determined in advance using various statistics.
Source vector with real values is replaced by a vector binary values $0$ and $1$ ($false$ and $true$). The value of the $i$-th attribute $f_i$ is replaced by the binary vector $g_i$ where $g_{ij} = f_i > B_{ij}$. Finally, all $g_i$ are concatenated into one vector. Consecutively, in the resuling trees instead of comparing some real feature value with some threshold there will be just boolean check.

Number of baskets could by controlled by passing the appropriate parameter during learning procedure.

Also another important parameter should be marked. It is the speed of learning. The smaller it is, the smaller the contribution of each new tree. However, this has to be compensated by a large number of trees. The speed of learning can be selected by cross-validation process with the fixed number of trees.
Usually, more trees means higher quality.

\section{Decision Tensor}
One of the simplest ways to accelerate is the preliminary calculation. The feature vector obtained after binarization takes a limited number of values. Therefore, we can precompute the prediction for all such values.

Let $n$ be the number of features. Let us consider some model $Z$, trained by some algorithm, containing a set of trees $T$. For convenience, we assume the values of the features lie in thre interval $[0, 1]$.

Let $f_i$ be the set of basket boundaries for the $i$-th feature, which was used in $T$. Then using the planes $X_i = b_{ij}$ for all $i,j$, we can split the space $[0;1]$ int small hyperparallelepipeds. It is obvious that within each hyperparallelepiped the prediction of the model is constant.
In fact, the model $Z$ allows you to calculate the value at any point of our hypercube (or Tensor). However, if we pre-calculate the value in each piece, then we will accelerate the application of our model, reducin it only determining the desired piece. The asymptotic complexity of the application of the tree-based mode it $O(h\cdot t)$ where $h$ is the height of the trees, $t$ is the total number of trees. Asymptotic complexity of Tensor is $O(n\cdot log(b))$ where $n$ is the number of features and $b$ is the number of cuts per feature (this complexity is achieved by using binary search). It should be marked for small $b$ simple linear search could be a bit faster that binary. So, with a small number of features Tensor can be applied much faster than usual tree-based model.

However, Decision Tensor has large memory footpring. It needs $O(b^n)$ of memory and it may be too large for common computer. Therefore, the precomputed tensor can only be used for small $b$ and $n$.

In the general, the size of one tensor can be defined as follows:

@@

Consider the case for small n. We can get a small $b$ by two main ways: limit the number of cuts for each feature and train model with big enough tree count or severely limit the tree count, then the ottal number of really used cuts will be exactly $h\cdot t$ without taking into account repeated use of the same cuts.

The first method was used in [2]: there was trained a CatBoost with a small number of cuts and turned into a Decision Tensor.

The consumption of memory in one hypercube can be extremely huge provded that acceptable quality is achieved, so the natural continuation is the construction of several Decision Tensors.

\section{Tensor Similarity}
Introduce the metric of the similarity of two tensors. Let's define it as the difference between the sum of their sizes and the size of their union:

@@

Target metrics are $ Q $ (model quality), $ S $ (size), and $ t $ (time of application). However, we can assume that the application time is proportional to the number of decision tensors $ M $.
The input vector will need to describe the algorithm for obtaining trees and the algorithm for combining them into cubes. We denote it by $ v $ and by $V$ the feasible set of input vectors. Then we need to solve the following problem:

@@

Before we move to consider the contents of the vector $ X $, it should be noted that for practical application this problem can be solved using the constraint method. Usually in practice there are quite strict limitations on the size of the model and the available application time.

\section{Algorithm of combining a particular set of trees in a set of decision tensors}

Suppose we have $N$ trees and we want to combine them in $M$ of the decision tensors. The metric of optimality of the partition will be assumed to be occupied memory.
In general, this is not an easy task and it requires a separate study. We used the following heuristic approach:
\medskip
\renewcommand{\theenumi}{\arabic{enumi}}
\begin{enumerate}
\item Initialize $M$ tensors by placing in each one an arbitrarily chosen tree
\item Choose any of the remaining trees
\item Combine it with the most similar tensor
\item Repeat steps 2-3 until there are no unused trees left.
\end{enumerate}

@@

\medskip
\begin{itemize}
\item We fix a very rough partition with a small number of cubes and train an arbitrarily large number of trees.
\item The partitioning is done quite detailed, but at the same time we significantly reduce the number of trained trees, thereby reducing the number of actually used buckets.
\end{itemize}

The algorithm actively uses both these ideas.

\medskip
\renewcommand{\theenumi}{\arabic{enumi}}
\begin{enumerate}
\item Determine the desired number of tensors (M) based on time constraints
\item Train the CatBoost model with a limited number of buckets
\item Unite the trees from the obtained model into the first decisive tensor from the ensemble
\item Use random search to sort through the key parameters of CatBoost
\medskip
\begin{itemize}
\item Train the CatBoost model, using the previously trained model as the baseline
\item Transform into M-1 Decision Tensors
\item Remove ensembles that do not fit the required size limits.
\end{itemize}
\item We choose the ensemble with the best quality on an additional test set of data.
\end{enumerate}

This algorithm allows to find the optimal Pareto ensemble of Decision Tensors.

\section{Application of the algorithm on the dataset}

We have used the dataset and the model from [@ likhomanenko2015lhcb]. Number of features in the dataset is 10, the model was trained on 5000 trees with a number of the buckets 64. We compared our algorithm for constructing Decision Tensors Ensemble with trivial approach: training a small number of trees but with high value of learning speed.
For this we divided the data set into three parts: teaching (50%), test (25%) and validation (25%). We trained the models on the training sample, then using the test we selected the best quality models under given constraints.

We have limited the number of tensor by 6 and the total size of models by one gigabyte, believing every prediction requires 4 bytes (float). We looked through only the key parameters of CatBoost about the number of trees and the speed of learning.

The area under the ROC curve where False Positive Rate less than 0.05 was taken as a metric. For convenience, we have adjusted this metric so that its maximum value (with an ideal model) is equal to 1.

@@

We compared the CatBoost model with 5000 trees ("best"), the fast CatBoost models for 1-125 trees ("fast") and the Decision Tensors Ensembles with different memory limits and different tensors numbers.

Chart 1 shows the dependence of quality on the number of lookups. Chart 2 shows the dependence of quality on the total time of application at the full validation dataset. Table 3 shows properties of the models trained in the process.

@@

However, the difference in working time is not so significant - 2.5 times and 50 times, respectively. This is due to the principles of the cache and the specialty of our benchmark. Since the Decision Tensor takes up a very large amount of memory and does not fit into the cache because of this, almost every lookup results in reading from the RAM. CatBoost models are small enough, because of what can be entirely placed in the cache. Since our benchmark consists only of model applications, CatBoost models are not forced out of the cache. In real applications, the CatBoost model will most likely not own the cache alone and will be even slower because of this. And Decision Tensor Ensemble almost do not depend on the cache size.

@@

Decision Tensor Ensbles allow you to simplify the application of complex Boosting Decision Trees algorithms (such as CatBoost) in real-time systems without the ability to use batching application.
Considered algorithm for building ensembles allows you to exchange consumed memory for the speed of application with a small loss in quality.
In addition, the use of the Ensemble can significantly improve the quality in comparison with the use of a single Decision Tensor.

>>>>

For the most of the ML algorithms, improving the quality of the model usually leads to a higher levels of resource consumption. For example, with the use tree-based boosting algorithms, the increase of the number of trees from tens to thousands usually leads to the significant increase in the quality and at the same time the model size and inference time also amplify. The higher demands for the hardware resources for the more accurate models are usually solved by various technical optimizations model storage mechanism. For example vectoriation of the tree weights storage can be used. Another approach is the modification of the training process that makes it possible to obtain models with certain properties that make it possible to signigicantly accelerate their application without maintaining the required level of quality.

In this paper, we consider the modification of the training process of a predictive models by decision tree-based algorithm CatBoost [1]. Such modification will accelerate the prediction time of the model while maintaining the required level of the quality.

@@

CatBoost is the implementation of a tree-based gradient boosting in which so-caled oblivious decision trees are used. The values of each feature $i$ are discretized into baskets using the boundaries $B_i$. The boundaries themselves are determined in advance using various heuristic.
Original feature vector with real values is replaced by a vector binary values $0$ and $1$ ($false$ and $true$). The value of the $i$-th attribute $f_i$ is replaced by a binary vector $g_i$ where $g_{ij} = f_i > B_{ij}$. Finally, all $g_i$ are concatenated into one vector. Consecutively, the resuling decision will make bunch of boolean checks instead of comparing some real feature value with the thresholds.

Number of baskets used for the discretization could be controlled by specifying the appropriate parameter during the training procedure.

Let's consider another important training characteristic, namely the learning rate. The smaller it is, the smaller the contribution of each next tree in the ensemble. However, this has to be compensated by a large number of trees. The learning rate can be chosen by the cross-validation process given the fixed number of trees. Usually, the more trees the higher the model quality.

\section{Decision Tensor}
One of the simplest ways to accelerate computational performance is the preliminary calculation. The feature vector obtained after the discretization takes a limited number of values. Therefore, we can precompute the predictions for all possible values combination.

Let $n$ be the number of features. Let us consider some model $Z$, trained by some algorithm, containing a set of trees $T$. For the convenience, we assume the values of every feature belongs to the interval $[0, 1]$.

Let $f_i$ be the set of basket boundaries for the $i$-th feature, which was used in $T$. Then we can split the interval $[0;1]$ into small hyper-parallelepipeds. The prediction of the model is constant within each hyperparallelepiped. In fact, the model $Z$ allows you to calculate the value at any point of our hypercube that we call Tensor. However, if we pre-calculate the value for each piece, then we will accelerate the application of our model, effectively reducing it to the time requited for the determination of the desired piece. The asymptotic complexity of the application of the tree-based mode it $O(h\cdot t)$ where $h$ is the height of the trees, $t$ is the total number of the trees. Asymptotic complexity of Tensor is $O(n\cdot log(b))$ where $n$ is the number of the features and $b$ is the number of cuts per feature (this complexity is achieved by using binary search). (It should be noted for small $b$ simple linear search could be a bit faster that binary). So, with a small number of features the Tensor-based prediction can be computed much faster than usual tree-based model.

However, Decision Tensor has larger memory footprint. It needs $O(b^n)$ bytes of memory and it may be too large for common computer. Therefore, the precomputed tensor can only be used for small $b$ and $n$.

In the general, the size of one tensor can be estimated as follows:

@@

Consider the case for small $n$. We can get a small $b$ in two main ways: limit the number of cuts for each feature and train the model with big enough tree count or severely limit the tree count, then the total number of effective cuts will be exactly $h\cdot t$ without taking into account repeated use of the same cuts.

The first method was used in [2]: there was a trained CatBoost model with a small number of cuts that has been precomputed into a Decision Tensor.

The consumption of memory for a single hypercube can be extremely huge provded that acceptable quality is achieved, so the natural extension of the idea is the construction of Decision Tensors ensemble.

\section{Tensor Similarity}
Let's introduce the metric for the similarity of two tensors. Let's define it as the difference between the sum of their sizes and the size of their union:

@@

The target metrics are $ Q $ (model quality), $ S $ (size), and $ t $ (time of inference). However, we can assume that the inference time is proportional to the number of decision tensors $ M $.
The input feature vectors will define the algorithm for obtaining the trees and the algorithm for combining them into the Tensors. We denote single vector by $ v $ and let $V$ be the set of all input vectors. Then we need to solve the following problem:

@@

It should be noted that for practical application this problem can be solved using the constrained optimization method. But in practice there are quite strict limitations on the size of the model and the available application time (TODO - what did you mean?).

\section{Algorithm for combining set of Decision Trees to a set of Decision Tensors}

Suppose we have $N$ trees and we want to combine them in $M$ of the Decision Tensors. The efficiency metric of the partition is the result Tensor set memory footprint.
In general, this is not an easy task and it requires a separate study. We used the following heuristic approach:
\medskip
\renewcommand{\theenumi}{\arabic{enumi}}
\begin{enumerate}
\item Initialize $M$ Tensors by placing in each one an arbitrarily chosen tree
\item Choose any of the remaining trees
\item Combine it with the most similar tensor
\item Repeat steps 2-3 until there are no unused trees left.
\end{enumerate}
Such reduction of the set of trees to a set of Tensors we call partitioning. 
(TODO: what is rough and what is detailed partition?)

@@

\medskip
\begin{itemize}
\item We fix a very rough partition by a small number of Tensors and train an arbitrarily large number of trees.
\item The partitioning is done quite detailed, but at the same time we significantly reduce the number of trained trees, thereby reducing the number of actually used buckets.
\end{itemize}

The algorithm actively uses both of these ideas.

\medskip
\renewcommand{\theenumi}{\arabic{enumi}}
\begin{enumerate}
\item Determine the desired number of Tensors (M) based on time constraints
\item Train the CatBoost model with a limited number of buckets
\item Merge the trees from the obtained model into the first decisive tensor from the ensemble
\item Chose another set of CatBoost hyperparameters using Random Search
\medskip
\begin{itemize}
\item Train the CatBoost model, using the previously trained model as the baseline
\item Transform it into M-1 Decision Tensors
\item Remove ensembles that do not fit the required size limits (TODO: specify limits at the beginning?).
\end{itemize}
\item We choose the ensemble with the best quality on an additional data sample.
\end{enumerate}

This algorithm allows to find the Pareto-optimal ensemble of Decision Tensors.

\section{Real problem illustration}

We have used the dataset and the model from [@ likhomanenko2015lhcb]. Number of features in the dataset is 10, the model was trained on 5000 trees with a number of the buckets 64. We compared our algorithm for constructing Decision Tensors Ensemble with trivial approach: training a small number of trees but with high value of the training speed.
For this we divided the data set into three parts: training (50\%), test (25\%) and validation (25\%). We trained the models on the training sample, then used the test for the selection of the best models under given constraints.

We have limited the number of tensor by 6 and the total size of models by one gigabyte, so every prediction requires 4 bytes (float). We have scanned only through the key CatBoost hyperparameters: number of trees and the learing rate.

The figure of merit we've taken also from [@_likhomanenko2015lhcb], which is the area under the ROC curve where False Positive Rate less than 0.05. For the convenience, we have normalized this metric so that its maximum value (with an ideal model) is equal to 1.

@@

We compared the CatBoost model with 5000 trees ("best"), the fast CatBoost models for 1-125 trees ("fast") and the Decision Tensors Ensembles (DTE) with different memory limits and different tensors numbers.

Figure 1 shows the dependence of the quality on the number of lookups. Chart 2 shows the dependence of the quality vs the total inference time at the full validation dataset. Table 3 shows properties of the models trained in the process.

@@

However, the difference in working time is not so significant - 2.5 times and 50 times, respectively. This is due to the principles of the CPU caching and the size of our dataset. Since the Decision Tensor takes up a very large amount of memory and does not fit into the CPU cache because, almost every lookup results in reading from the RAM. CatBoost models are small enough to fit into the cache. Since our benchmark consists only of model applications, CatBoost models are not moved out of the cache. In real applications, the CatBoost model will most likely not sit in the cache alone and will be even slower because of this. On the contrary the performance of Decision Tensor Ensemble will not degrade due to that fact.

@@

Decision Tensor Ensbles allow you to speed up the inference by complex Boosting Decision Trees algorithms (such as CatBoost) in real-time systems. Considered algorithm for building Tensor ensembles allows for additional tradeoff between memory and the inference speed without significant loss of the model quality.
In addition, the use of the Ensemble can significantly improve the model quality in comparison with a single Decision Tensor.
